Gamma-Ray Burst Afterglows In The Multi-Messenger Era: Numerical Models and Closure Relations

### 2020-05-13 refereeResponse.txt

We thank the referee for clearing up the history of the subject, and have amended the introduction as necessary.

-- Expanded discussion and added references concerning history of structured jets in paragraphs 5 and 6 of Section 1.

7. Figure 1 presented some structured jet models from numerical simulations. Figure 4 of this paper
https://ui.adsabs.harvard.edu/abs/2019ApJ...877L..40G/abstract
provide more structures. In particular, the shape of E(\theta) depends on the waiting time before the jet is launched. For a delayed launch, one has essentially a two-component jet structure (jet + cocoon). For a prompt launch (negligible delay), the jet penetrates through the ejecta quickly and develop a structured jet outside the ejecta. The jet structure is one-component, similar to the Gaussian profile promoted in this paper. This may be discussed in the paper.

-- Added M0.1 model from Geng+ to Figure 1, amended the caption, and added a description to the second paragraph of Section 2.

8. The most complete "closure relations" for GRB afterglow models was presented in a review article
https://ui.adsabs.harvard.edu/abs/2013NewAR..57..141G/abstract.
This may be referenced when closure relations are discussed.

-- Added a reference and sentence description of the paper to the first paragraph of Section 4.

Other updates to the text:

@@

-- Added a sentence to the end of the fourth-to-last paragraph of Section 5 explaining the origin of the constraining power on theta_obs and theta_c.

-- Added a paragraph to section 6.1 discussing the role of gravitational wave data in constraining theta_obs and theta_c.

-- Added a couple sentences to the 6th paragraph of Section 6.1 referencing using Luminosity functions and rates to constrain jet structure, citing recent Beniamini+ works

-- Added Mooley 2018c to references for 170817 post-jet break slope in Section 1 and 5

-- Added paragraph discussing Hajela+ 2019 analysis to Section 6.2.

### 2020-05-13 structuredJets.tex
Finished referee response?

The binary neutron star merger event \gwbns{}, followed quickly by the short gamma-ray burst \grbbns{}, provided a new view on gamma-ray burst (GRB) afterglows \citep{Abbott:2017ab, Abbott:2017ac}.  Unlike typical afterglows, which begin bright and decay from detection with a ~week timescale, \grbbns{}'s non-thermal emission was undetectable until the first observation of X-rays nine days after the GRB \citep{Troja:2017aa}.  At this point the afterglow began steadily increasing in brightness at all wavelengths for $\sim160$ days \citep{Haggard:2017aa, Hallinan:2017aa, DAvanzo:2018aa,  Lyman:2018aa, Margutti:2018aa, Mooley:2018aa, Troja:2018aa, Troja:2019ab}.  Very long baseline (VLBI) radio imagery over this period identified a radio core with an apparent superluminal motion, indicating the emitting surface was moving at relativistically at an oblique angle towards the Earth \citep{Mooley:2018ab}. The emission peaked at 164 days after the burst and proceeded to sharply decay at a rate $t^{-2.2}$ commensurate with other GRB afterglows \citep{Alexander:2018aa, Fong:2019aa, Lamb:2019aa, Troja:2019ab}.

The lack of early emission, slow rising light curve, apparent motion of the radio centroid, and sharp decline post-peak are all consistent with emission from a \emph{structured jet}, a collimated blast wave with a non-trivial angular distribution of energy, viewed a moderate angle away from the jet axis \citep{Lamb:2017aa, Alexander:2018aa, Hotokezaka:2018aa, Wu:2018aa, Xie:2018aa,  Ghirlanda:2019aa, Fong:2019aa, Lamb:2019aa, Troja:2019aa}.  The combined effects of a moderate viewing angle and angular structure in the jet produce a light curve significantly different from the standard on-axis uniform ``top-hat'' jet used extensively in GRB afterglow analysis. Being sources at cosmological distances, GRBs must typically be observed nearly on-axis or at least within the original opening angle of the jet \citep{Ryan:2015aa}. At these small viewing angles lateral structure plays a sub-dominant role and the ubiquitous top hat jet has been a sufficient model for most studies.  Since the gravitational wave signal of a binary neutron star merger is nearly isotropic, it is expected the majority of future GW-GRBs will be viewed at significant inclination and may have similarly peculiar light curves as \grbbns{}.

@@

Numerical simulations have revealed a variety of jet angular energy distributions, often containing an energetic core with power law tails.  Figure \ref{fig:jetStruct} shows a collection of jet energy distributions from the literature \citep{Aloy:2005aa, Mizuta:2009aa, Duffell:2013aa, Lazzati:2017aa, Margutti:2018aa}.  The \citet{Aloy:2005aa} model (B01 in the paper) was launched as a top hat jet into an accretion torus featuring a narrow underdense funnel region and produced a final energy profile with a relatively sharp edge, although only material with Lorentz factor greater than 100 was included in the $\Eiso$ calculation.  The \citet{Margutti:2018aa} model used a powerful engine with Gaussian injection profile into a standard BNS merger cloud, resulting in a blunt jet with less energetic wings.  The \citet{Mizuta:2009aa} model (their HE16N simulation) is of a collapsar jet, injected with a $5^\circ$ opening angle into a collapsing massive star density profile.  A large degree of interaction with the stellar envelope produces the energetic power law wings.  The \citet{Lazzati:2017aa} model is meant to emulate a BNS merger and injects a top hat jet into a spherically symmetric ejecta wind, also resulting in a large amount of interaction and energetic power law wings.  The \citet{Duffell:2013aa} model is an analytic model of a fireball, spherically symmetric in its rest frame, boosted at a bulk Lorentz factor. It also produces a narrow core with power law wings.

@@

\begin{figure}
	\includegraphics[width=\columnwidth]{figs/jetStructureComparison.pdf}
	\caption{Lateral profiles of isotropic equivalent energy $\Eiso$ as a function of angle from the jet axis $\theta$, individually rescaled to group similar profile shapes.  The thick grey lines show fiducial profiles with simple analytic expressions (Equations \eqref{eq:gauss} and \eqref{eq:pl}) while the thin colored lines show results from numerical simulations and analytic models chosen from the literature.  The blue line is the B1 numerical simulation of \cite{Aloy:2005aa}, orange the HE16N numerical simulation of \cite{Mizuta:2009aa}, green an analytic ``boosted fireball'' model with $\gamma_{\rm{B}}=10$ and $\eta_0=3$ from \cite{Duffell:2013aa}, red the numerical simulation from \cite{Lazzati:2017aa} (L17), and purple the numerical simulation from \cite{Margutti:2018aa} (M18).  The thick solid grey line is a Gaussian profile (Equation \eqref{eq:gauss}) with $E_0 = 3\times 10^{52}$ erg, $\thC = 6^\circ$, and $\thW=12^\circ$ and the thick dashed grey line is a power law profile (Equation \eqref{eq:pl}) with $E_0 = 10^{52}$ erg, $\thC = 2^\circ$, $\thW=20^\circ$, and $b=4.5$.  The Gaussian and power law profiles can emulate the basic properties of the energy profiles found in the literature. \label{fig:jetStruct}}
\end{figure}

@@

Afterglow light curves, even when viewed off axis, can typically be described as a piecewise set of power law segments $F_\nu \propto \tobs^\alpha \nuobs^\beta$, each denoting particular phases of evolution of the blast wave and its emitting region.  Particular phases have been identified empirically \citep{Nousek:2006aa} and theoretically \citep{Zhang:2006aa}.   Within a particular phase, the slopes $\alpha$ and $\beta$ are often related through a \emph{closure relation}: an equality determining $\alpha$ from $\beta$ or vice versa.  Phases change at \emph{break times} when the temporal slope $\alpha$ (and potentially the spectral slope $\beta$) transition rapidly from the previous phase to the next.

@@

Figure \ref{fig:cornerGaussian} shows the observations in radio, optical, and X-ray bands of the \gwbns{} afterglow.  No afterglow emission was detected until an X-ray detection by the Chandra observatory nine days after the GW trigger \cite{Troja:2017aa}.  Further radio, optical, and X-ray observations showed a steady rising light curve with constant power law slope until a peak near day 164 \citep{Haggard:2017aa, Hallinan:2017aa, DAvanzo:2018aa,  Lyman:2018aa, Margutti:2018aa, Mooley:2018aa, Troja:2018aa, Troja:2019ab}.  At this point the light curve turned over and began a steep decay that continues to the present \citep{Alexander:2018aa, Fong:2019aa, Lamb:2019aa, Troja:2019ab}.

@@

In the future, it may be possible to attack the question of determining $E(\theta)$ at the population level.  Figure \ref{fig:gPop} shows the structure parameter $g$ as a function of $\thobs/\thC$ for several jet models, as well as the inferred value from \gwbns{}.  If all short GRB jets share a jet profile shape, they should trace out a single curve in this space.  As more misaligned afterglows are observed, measurements of $g$ can populate this diagram and potentially rule out models (such as potentially the $b=2$ power law).  Model-independent constraints on $\thobs/\thC$ will greatly aid this procedure.  Moreover, each structure model should predict a different observed distribution of $g$, depending on the shape of the $g(\thobs/\thC)$ curve and the detectability of each burst.  With sufficient observations of $g$ alone and understanding of the observational biases, it may be possible to infer the underlying $g(\thobs/\thC)$ curve and $E(\theta)$ itself.

@@

Many groups have used structured jet models to analyze the afterglow of \gwbns{}, among the most recent with comparable models are \citet{Hotokezaka:2018aa, Ghirlanda:2019aa, Lamb:2019aa, Wu:2018aa}.  Of these, \citet{Hotokezaka:2018aa, Ghirlanda:2019aa, Lamb:2019aa} use semi-analytic methods similar to the present work to construct afterglow light curves, while \citet{Wu:2018aa} utilizes analytic scaling relations and a template bank constructed from numerical simulations (following the conceptual approach introduced by \citet{van-Eerten:2012ac}).  Additionally, \citet{Hotokezaka:2018aa, Ghirlanda:2019aa} include constraints from the VLBI measurements of the \gwbns{} radio centroid's apparent superluminal motion.

@@

>>>>

The binary neutron star merger event \gwbns{}, followed quickly by the short gamma-ray burst \grbbns{}, provided a new view on gamma-ray burst (GRB) afterglows \citep{Abbott:2017ab, Abbott:2017ac}.  Unlike typical afterglows, which begin bright and decay from detection with a ~week timescale, \grbbns{}'s non-thermal emission was undetectable until the first observation of X-rays nine days after the GRB \citep{Troja:2017aa}.  At this point the afterglow began steadily increasing in brightness at all wavelengths for $\sim160$ days \citep{Haggard:2017aa, Hallinan:2017aa, DAvanzo:2018aa,  Lyman:2018aa, Margutti:2018aa, Mooley:2018aa, Troja:2018aa, Troja:2019ab}.  Very long baseline (VLBI) radio imagery over this period identified a radio core with an apparent superluminal motion, indicating the emitting surface was moving at relativistically at an oblique angle towards the Earth \citep{Mooley:2018ab}. The emission peaked at 164 days after the burst and proceeded to sharply decay at a rate $t^{-2.2}$ commensurate with other GRB afterglows \citep{Alexander:2018aa, Fong:2019aa, Lamb:2019aa, Mooley:2018ac, Troja:2019ab}.

The lack of early emission, slow rising light curve, apparent motion of the radio centroid, and sharp decline post-peak are all consistent with emission from a \emph{structured jet}, a collimated blast wave with a non-trivial angular distribution of energy, viewed a moderate angle away from the jet axis \citep{Lamb:2017aa, Alexander:2018aa, Hotokezaka:2018aa, Wu:2018aa, Xie:2018aa,  Ghirlanda:2019aa, Fong:2019aa, Lamb:2019aa, Troja:2019aa, Hajela:2019aa}.  The combined effects of a moderate viewing angle and angular structure in the jet produce a light curve significantly different from the standard on-axis uniform ``top-hat'' jet used extensively in GRB afterglow analysis. Being sources at cosmological distances, GRBs must typically be observed nearly on-axis or at least within the original opening angle of the jet \citep{Ryan:2015aa}. At these small viewing angles lateral structure plays a sub-dominant role and the ubiquitous top hat jet has been a sufficient model for most studies.  Since the gravitational wave signal of a binary neutron star merger is nearly isotropic, it is expected the majority of future GW-GRBs will be viewed at significant inclination and may have similarly peculiar light curves as \grbbns{}.

@@

Numerical simulations have revealed a variety of jet angular energy distributions, often containing an energetic core with power law tails.  Figure \ref{fig:jetStruct} shows a collection of jet energy distributions from the literature \citep{Aloy:2005aa, Mizuta:2009aa, Duffell:2013aa, Lazzati:2017aa, Margutti:2018aa, Geng:2019aa}.  The \citet{Aloy:2005aa} model (B01 in the paper) was launched as a top hat jet into an accretion torus featuring a narrow underdense funnel region and produced a final energy profile with a relatively sharp edge, although only material with Lorentz factor greater than 100 was included in the $\Eiso$ calculation.  The \citet{Margutti:2018aa} model used a powerful engine with Gaussian injection profile into a standard BNS merger cloud, resulting in a blunt jet with less energetic wings.  The \citet{Mizuta:2009aa} model (their HE16N simulation) is of a collapsar jet, injected with a $5^\circ$ opening angle into a collapsing massive star density profile.  A large degree of interaction with the stellar envelope produces the energetic power law wings.  The \citet{Lazzati:2017aa} model is meant to emulate a BNS merger and injects a top hat jet into a spherically symmetric ejecta wind, also resulting in a large amount of interaction and energetic power law wings.  {\bf The \citet{Geng:2019aa} model launches a relativistic magnetized jet into a fiducial BNS merger ejecta cloud with varying delay times.  In this model the jet was launched 0.1 s after the ejecta, resulting in limited interaction before the jet broke out of the expanding cloud. } The \citet{Duffell:2013aa} model is an analytic model of a fireball, spherically symmetric in its rest frame, boosted at a bulk Lorentz factor. It also produces a narrow core with power law wings.

@@

\begin{figure}
	\includegraphics[width=\columnwidth]{figs/jetStructureComparison.pdf}
	\caption{Lateral profiles of isotropic equivalent energy $\Eiso$ as a function of angle from the jet axis $\theta$, individually rescaled to group similar profile shapes.  The thick grey lines show fiducial profiles with simple analytic expressions (Equations \eqref{eq:gauss} and \eqref{eq:pl}) while the thin colored lines show results from numerical simulations and analytic models chosen from the literature.  The blue line is the B1 numerical simulation of \cite{Aloy:2005aa}, orange the HE16N numerical simulation of \cite{Mizuta:2009aa}, green an analytic ``boosted fireball'' model with $\gamma_{\rm{B}}=10$ and $\eta_0=3$ from \cite{Duffell:2013aa}, red the numerical simulation from \cite{Lazzati:2017aa} (L17), purple the numerical simulation from \cite{Margutti:2018aa} (M18), and brown the M0.1 numerical simulation of \cite{Geng:2019aa}.  The thick solid grey line is a Gaussian profile (Equation \eqref{eq:gauss}) with $E_0 = 3$, $\thC = 6^\circ$, and $\thW=12^\circ$ and the thick dashed grey line is a power law profile (Equation \eqref{eq:pl}) with $E_0 = 1$, $\thC = 2^\circ$, $\thW=20^\circ$, and $b=4.5$.  The Gaussian and power law profiles can emulate the basic properties of the energy profiles found in the literature. \label{fig:jetStruct}}
\end{figure}

@@

Afterglow light curves, even when viewed off axis, can typically be described as a piecewise set of power law segments $F_\nu \propto \tobs^\alpha \nuobs^\beta$, each denoting particular phases of evolution of the blast wave and its emitting region.  Particular phases have been identified empirically \citep{Nousek:2006aa} and theoretically \citep[e.g.][]{Zhang:2006aa}.   Within a particular phase, the slopes $\alpha$ and $\beta$ are often related through a \emph{closure relation}: an equality determining $\alpha$ from $\beta$ or vice versa.  Phases change at \emph{break times} when the temporal slope $\alpha$ (and potentially the spectral slope $\beta$) transition rapidly from the previous phase to the next. {\bf A review and comprehensive collection of known GRB afterglow phases and closure relations is given in \citet{Gao:2013aa}.}

@@

Figure \ref{fig:cornerGaussian} shows the observations in radio, optical, and X-ray bands of the \gwbns{} afterglow.  No afterglow emission was detected until an X-ray detection by the Chandra observatory nine days after the GW trigger \cite{Troja:2017aa}.  Further radio, optical, and X-ray observations showed a steady rising light curve with constant power law slope until a peak near day 164 \citep{Haggard:2017aa, Hallinan:2017aa, DAvanzo:2018aa,  Lyman:2018aa, Margutti:2018aa, Mooley:2018aa, Troja:2018aa, Troja:2019ab}.  At this point the light curve turned over and began a steep decay that continues to the present \citep{Alexander:2018aa, Fong:2019aa, Lamb:2019aa, Mooley:2018ac, Troja:2019ab}.

@@

 In the future, it may be possible to attack the question of determining $E(\theta)$ at the population level. {\bf For instance, since GRB jets have isotropically distributed orientations, the jet structure profile will imprint itself on the GRB luminosity function.  If the observational biases and selection effects are understood, one can use the observed luminosity function or event rate to constrain structure models \citep{Beniamini:2019aa, Beniamini:2019ab}.  }

One possible new approach is to make direct use of the structured phase closure relations.  Figure \ref{fig:gPop} shows the structure parameter $g$ as a function of $\thobs/\thC$ for several jet models, as well as the inferred value from \gwbns{}.  If all short GRB jets share a jet profile shape, they should trace out a single curve in this space.  As more misaligned afterglows are observed, measurements of $g$ can populate this diagram and potentially rule out models (such as potentially the $b=2$ power law).  Model-independent constraints on $\thobs/\thC$ will greatly aid this procedure.  Moreover, each structure model should predict a different observed distribution of $g$, depending on the shape of the $g(\thobs/\thC)$ curve and the detectability of each burst.  With sufficient observations of $g$ alone and understanding of the observational biases, it may be possible to infer the underlying $g(\thobs/\thC)$ curve and $E(\theta)$ itself.

@@

Many groups have used structured jet models to analyze the afterglow of \gwbns{}, among the most recent with comparable models are \citet{Hotokezaka:2018aa, Ghirlanda:2019aa, Lamb:2019aa, Wu:2018aa, Hajela:2019aa}.  Of these, \citet{Hotokezaka:2018aa, Ghirlanda:2019aa, Lamb:2019aa} use semi-analytic methods similar to the present work to construct afterglow light curves, while \citet{Wu:2018aa} and \citet{Hajela:2019aa} utilize analytic scaling relations and a template bank constructed from numerical simulations (following the conceptual approach introduced by \citet{van-Eerten:2012ac}).  Additionally, \citet{Hotokezaka:2018aa, Ghirlanda:2019aa} include constraints from the VLBI measurements of the \gwbns{} radio centroid's apparent superluminal motion.

@@

{\bf \citet{Hajela:2019aa} uses the same model as \citet{Wu:2018aa} with an updated afterglow dataset and a new upper limit on $n_0$ of $9.6\times 10^{-3}$ cm$^{-3}$ computed from a measure of the diffuse X-ray emission in the host galaxy.  They find $\thobs = 0.53^{+0.07}_{-0.06}$ rad, larger than \citet{Wu:2018aa}, also using only afterglow data.  Their posterior distributions also display a strong relation $\gamma_B \thobs \approx 5$,  and values of $E_{tot}$ (their $E_0$), $n_0$, $\epse$, and $\epsB$ consistent with other studies. }
